### 2018-04-21 Manuscript/chapter4.tex
Final changes before addition to thesis

Although there is growing interest in how viruses and diet quality affect the health and sustainability of honeybees, as well as a recognition that such factors might operate interactively, there are only a small number of experimental studies thus far directed toward elucidating the interactive effects of these two factors in honeybees (\citealt{intNV}, \citealt{intNV2}, \citealt{intNV3}). We recently used laboratory cages and nucleus hive experiments to investigate the health effects of these two factors, and our results show a significant interaction between diet quality and virus infection. Specifically, high quality pollen is able to mitigate virus-induced mortality to the level of diverse, polyfloral pollen (\citealt{adamInt}).

Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms by which high quality diets protect bees from virus-induced mortality. For example, it is not known whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (resilience). Transcriptomics is one means to achieve this goal. Transcriptomic analysis can help us identify 1) the genomic scale of transcriptomic response to diet and virus infection, 2) whether these factors interact in an additive or synergistic way on transcriptome function, and 3) the types of pathways affected by diet quality and viral infection. This information, heretofore lacking in the literature, can help us better understand how good nutrition may be able to serve as a "buffer" against other stressors (\citealt{AdamTothReview}). As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects (\citealt{alaux2}) and IAPV infection effects (\citealt{galbraith}) in honeybees. As far as we know, there are few to no studies investigating honeybee gene expression patterns specifically related to monofloral diets, and few to no studies investigating honeybee gene expression patterns related to the interaction effects of diet in any broad sense and viral innoculation in any broad sense.

In this study, we examine how monofloral diets and viral innoculation influence gene expression patterns in honeybees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure). We conduct RNA-sequencing analysis on a randomly selected subset of the honeybees we used in our previous study (as is further described in our methods section). We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the interactive effect of the two factors on gene expression patterns.

@@

Consequently, for our nutrition factor, we examined two monofloral pollen diets, Rockrose (Cistus) and Chestnut (Chestnut). Rockrose pollen is generally considered less nutritious than Chestnut pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron (\citealt{DiPasquale}, \citealt{adamInt}). For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV. This experimental design resulted in four treatment groups (Rockrose pollen without IAPV exposure, Chestnut pollen without IAPV exposure, Rockrose pollen with IAPV exposure, and Chestnut pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honeybees.

@@

\subsection{Detecting resilience versus resistance}

@@

@@

Mortality rates of honeybees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df=3, 55; F=10.07; p<2.18e-05). The effect of virus treatment (mixed model ANOVA, df=1, 55; F=24.343; p<7.84e-06) and diet treatment (mixed model ANOVA, df=1, 55; F=5.796; p<0.0194) were significant, but the interaction between the two factors (mixed model ANOVA, df=1, 55; F=0.062, p=0.8039) was not significant. The virus treatment was significant: For a given diet, honeybees exposed to the virus showed significantly higher mortality rate than honeybees not exposed to the virus (Tukey HSD, p<0.05). In comparing mortality levels based on pairwise comparisons, we found that bees fed Rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls. However, bees fed Castanea pollen had no significant difference in mortality between virus infected and control groups. These results suggest the high quality Castanea diet can ``rescue'' virus induced mortality (Figure \ref{fig:mortality}A).

IAPV titer volumes of honeybees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df=3, 34; F=6.096; p<0.00196). The effect of virus treatment (mixed model ANOVA, df=1, 34; F=15.686; p<0.000362) was significant, but the diet treatment (mixed model ANOVA, df=1, 34; F=1.898; p>0.05) and the interaction between the two factors (mixed model ANOVA, df=1, 34; F=0.702, p>0.05) were not significant. Honeybees that were infected with the virus and fed a poor-quality Rockrose diet showed significant increases in IAPV titer volumes compared to honeybees that were not infected with the virus regardless of their diet quality (Tukey HSD, p<0.05). Overall, we interpreted this effect to mean that Rockrose pollen could not ``rescue'' high virus titers resulting from the inoculation treatment, whereas Castanea pollen could (Figure \ref{fig:mortality}B).

@@

We observed a substantially larger number of DEGs in our diet main effect (n = 1914) than in our virus main effect (n = 43) (Table \ref{tbl:mainEffectDEGs}A and B). In the diet factor, there were more Chestnut-upregulated DEGs (n = 1033) than Rockrose-upregulated DEGs (n = 881). In the virus factor, there were more virus-upregulated DEGs (n = 38) than control-upregulated DEGs (n = 5). While these reported DEGs numbers are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Table \ref{tbl:mainEffectDEGs}A and B).

GO analysis of the Chestnut-upregulated DEGs revealed the following enriched categories (Benjamini correction < 0.05): Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Table \ref{tbl:ChestnutPathways}). GO analysis of the Rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Table \ref{tbl:RockrosePathways}).

@@

The number of DEGs across the six treatment pairings between the diet and virus factor ranged from 0 to 941 (Table \ref{tbl:pairDEGs}). Some of the trends observed in the main effect comparisons persisted: The diet level appeared to have greater influence on the number of DEGs than the virus level. Across every pair comparing the Chestnut and Rockrose levels, regardless of the virus level, the number of Chestnut-upregulated DEGs was higher than the number of Rockrose-upregulated DEGs (Table \ref{tbl:pairDEGs} C, D, E, F). For the pairs in which the diet level was controlled, the virus-exposed treatment showed equal to or more DEGs than the control treatment (Table \ref{tbl:pairDEGs} A, B). There were no DEGs between the treatment pair controlling for the control level of the virus effect (Table \ref{tbl:pairDEGs} A). These trends were observed for all three pipelines used (DESeq2, edgeR, and limma).

@@

We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset. Basic MDS plots were constructed with the DESeq2 analysis pipeline, and we could immediately determine that the Galbraith dataset may better separate the infected and uninfected honeybees better than our dataset (Figure \ref{fig:mdsPlots}). We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the MDS plots. However, through this automatically-generated plot, we can only visualize information at the sample level. Wanting to learn more about the data at the gene level, we continued with additional visualization techniques.

@@

We see that the 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays (Figure \ref{fig:pcpGalbraith}). We do see that the first replicate of the virus group appears somewhat inconsistent with the other virus replicates in Cluster 2, confirming that this trend in the data that we saw in the MDS plot carried through into the DEG results. In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure \ref{fig:pcpRutterVirus}). The replicates appear somewhat inconsistent in their esimated expression levels and there is not always such a large difference between treatment groups. We see a similar finding when we also examine a larger subset of 1,914 diet-related DEGs from our study.

We also used litre plots to examine the structure of individual DEGs: We see that indeed the individual DEGs from our data (Figure \ref{fig:litreClusterRutter}) show less consistent replications and less differences between the treatment groups compared to the individual DEGs from the Galbraith data (Figure \ref{fig:litreCluster1} and Figure \ref{fig:litreCluster2}). For the Galbraith data, we examined individual DEGs from the first cluster (Figure \ref{fig:litreCluster1}) and second cluster (Figure \ref{fig:litreCluster2}) because the second cluster was a bit less ideal due to its inconsistent first replicate of the treatment group.

Finally, we looked at scatterplot matrices to assess the DEGs. We created standardized scatterplot matrices for each of the four clusters (Figure \ref{fig:pcpGalbraith}) of the Galbriath data (Figures \ref{fig:GalbraithClust1SM}, \ref{fig:GalbraithClust2SM}, \ref{fig:GalbraithClust3SM}, and \ref{fig:GalbraithClust4SM}). We also created standardized scatterplot matrices for our data. However, as our dataset contained 24 samples, we would need to include 276 scatterplots in our matrix, which would be too numerous to allow for efficient visual assessment of the data. As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Figures \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, and \ref{fig:RutterSM4}). We can again confirm through these plots that the DEGs from the Galbraith data appeared more as expected: Deviating more from the \textit{x=y} line in the treatment scatterplots while staying close to the \textit{x=y} line in replicate scatterplots.

Despite the DEGs from the Galbraith dataset displaying the expected patterns more than those from our dataset, there was significant overlap in the DEGs between the two studies (Figure \ref{fig:GRVenn}).

@@

Within our ``resilience" gene ontologies, we found functions related to metabolism (such as carbohydrate metabolism, fructose metabolism, and chitin metabolism). However, we also discovered gene ontologies related to RNA polyerase II transcription and immune response (Figure \ref{fig:revigo}A). Within our ``resistance" gene ontologies, we found functions related to metabolism (such as carboyhydrate metabolism, chitin metabolism, and general metabolism).  (Figure \ref{fig:revigo}B).

@@

Challenges to honey bee health are a growing concern, in particular the combined, interactive effects of nutritional stress and pathogens (Dolezal and Toth 2018). In this study, we used RNA-sequencing to probe mechanisms underlying honey bee responses to two effects, diet quality and infection with the major virus of concern, IAPV. In general, we found a major nutritional transcriptomic response, with nearly 2,000 transcripts changing in response to diet quality (rockrose/poor diet versus chestnut/good diet). The majority of these genes were upregulated in response to high quality diet, and these genes were enriched for functions (Table \ref{tbl:ChestnutPathways}) such as nutrient signaling (insulin resistance) metabolism, and immune response (Notch signaling and JaK-STAT pathways). These data suggest high quality nutrition may allow bees to alter their metabolism, favoring investment of energy into innate immune responses.

Somewhat surprisingly, the transcriptomic response to virus infection in our experiment was fairly limited. We found only 43 transcripts to be differentially expressed, some with known immune functions (Table \ref{tbl:virusGenes}) such as argonaute-2 and a gene with similarity to MD-2 lipid recognition protein, as well as additional genes related to transcriptional regulation, and muscle contraction. The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figures \ref{fig:mdsPlots}B, \ref{fig:pcpRutterVirus}, \ref{fig:litreClusterRutter}, \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, and \ref{fig:RutterSM4}).

Given the noisy nature of our data, and our desire to hone in on genes with real expression differences, we compared our data to the Galbraith study (\citealt{galbraith}), which also examined bees response to viral infection. In contrast to our study, Galbraith et al. identified a large number of virus responsive trancripts, and generally had less noise in their data (Figures \ref{fig:mdsPlots}A, \ref{fig:pcpGalbraith}, \ref{fig:litreCluster2}, \ref{fig:litreCluster1}, \ref{fig:GalbraithClust1SM}, \ref{fig:GalbraithClust2SM}, \ref{fig:GalbraithClust3SM}, and \ref{fig:GalbraithClust4SM}). To identify the most reliable virus-responsive genes from our study, we looked for overlap in the DEGs associated with virus infection on both experiments. We found a large, statistically significant (p-value < 2.2e-16) overlap, with 26/38 (68\%) of virus-responsive DEGs from our study also showing response to virus infection in Galbraith et al. (Figure \ref{fig:GRVenn}). This result gives us confidence that, although noisy, we were able to uncover consistent, replicable gene expression responses to virus infection with our data.

The DESeq2 package comes with certain visualization options that are popular in RNA-sequencing analysis. One of these visualization is the multidimensional scaling (MDS) plot, which allows users to visualize the similarity between samples within a dataset. We could determine from this plot that indeed the Galbraith data may show more similarity between its replicates and differences between its treatments compared to our data (Figure \ref{fig:mdsPlots}). However, the MDS plot only shows us information at the sample level. We wanted to investigate how these differences in the signal:to:noise ratios of the datasets would affect the structure of any resulting DEGs. As a result, we also used three plotting techniques from the bigPint package: We investigated the 1,019 virus-related DEGs from the Galbraith dataset and the 43 virus-related DEGs from our dataset using parallel coordinate lines, litre plots, and scatterplot matrices. To prevent overlapping issues in our plots, we used a hierarchical clustering technique for the parallel coordinate lines to separate the set of DEGs into smaller groups. We also needed to examine four subsets of samples from the Galbraith dataset to make effective use of the scatterplot matrices. After these tailorizations, we determined that the same patterns we saw in the MDS plots regarding the entire dataset extended down the pipeline analysis into the DEG calls: Even the DEGs from the Galbraith datasaet showed more similarity between their replicates and differences between their treatments compared to those from our data. However, the 365 DEGs from the Galbraith data in Cluster 2 of Figure \ref{fig:pcpGalbraith} showed an inconsistent first replicate in the treatment group (``T.1''), which was something we observed in the MDS plot. This indicates that this feature also extended down the analysis pipeline into DEG calls.

One of the goals of this study was to use our RNA-seq data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of ``resistance'' or ``resilience''. We attempted to address this question through specific gene expression contrasts (Table \ref{tbl:contrasts}, Figure \ref{tbl:NCNRNCPathways}), accompanied by GO analysis of the associated gene lists. We found an approximately equal number of resistance (n = 122) and resilience (n = 125) related candidate genes, suggesting both processes may be playing significant roles in dietary buffering from pathogen infection. Resilience candidate genes had functions related to carbohydrate metabolism, chitin metabolism, immune response, and regulation of transcription. Resistance candidate genes had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion.

Overall, these data suggest complex transcriptomic responses to multiple stressors in honey bees. Diet has the potential for large and profound effects on transcriptional responses in honey bees, and differences in diet may set up the potential for both resistance and resilience to virus infection. Moreover, this study in general also demonstrated the possible benefits of examining multiple datasets to address inherently messy biological data. For instance, by verifying the subsantial overlap in our DEG lists to those obtained in another study that addressed a similar question but in a more controlled manner, we were able to feel more confident in our DEG calls.

\begin{table}[H]
  \includegraphics[width=0.65\textwidth]{Images/mainEffectDEGs}
  \caption{Number of DEGs across three analysis pipelines for (A) the diet effect in our study, (B) the virus main effect in our study, and (C) the virus main effect in the Galbraith study.}
  \label{tbl:mainEffectDEGs}
\end{table}

>>>>

Although there is growing interest in how viruses and diet quality affect the health and sustainability of honeybees, as well as a recognition that such factors might operate interactively, there are only a small number of experimental studies thus far directed toward elucidating the interactive effects of these two factors in honeybees (\citealt{intNV}, \citealt{intNV2}, \citealt{intNV3}). We recently used laboratory cages and nucleus hive experiments to investigate the health effects of these two factors, and our results show the importance of the combined effects of both diet quality and virus infection. Specifically, high quality pollen is able to mitigate virus-induced mortality to the level of diverse, polyfloral pollen (\citealt{adamInt}).

Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms by which high quality diets protect bees from virus-induced mortality. For example, it is not known whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (resilience). Transcriptomics is one means to better understand the mechanistic underpinnings of dietary and viral effects on honey bee health. Transcriptomic analysis can help us identify 1) the genomic scale of transcriptomic response to diet and virus infection, 2) whether these factors interact in an additive or synergistic way on transcriptome function, and 3) the types of pathways affected by diet quality and viral infection. This information, heretofore lacking in the literature, can help us better understand how good nutrition may be able to serve as a "buffer" against other stressors (\citealt{AdamTothReview}). As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects (\citealt{alaux2}) and IAPV infection effects (\citealt{galbraith}) in honeybees. As far as we know, there are few to no studies investigating honeybee gene expression patterns specifically related to monofloral diets, and few to no studies investigating honeybee gene expression patterns related to the combined effects of diet in any broad sense and viral innoculation in any broad sense.

In this study, we examine how monofloral diets and viral innoculation influence gene expression patterns in honeybees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure). We conduct RNA-sequencing analysis on a randomly selected subset of the honeybees we used in our previous study (as is further described in our methods section). We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the combined effect of the two factors on gene expression patterns.

@@

Consequently, for our nutrition factor, we examined two monofloral pollen diets, Rockrose (Cistus) and Castanea (Chestnut). Rockrose pollen is generally considered less nutritious than Chestnut pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron (\citealt{DiPasquale}, \citealt{adamInt}). For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV. This experimental design resulted in four treatment groups (Rockrose pollen without IAPV exposure, Chestnut pollen without IAPV exposure, Rockrose pollen with IAPV exposure, and Chestnut pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honeybees.

@@

\subsection{Probing resilience versus resistance}

@@

\begin{table}[H]
\begin{tabular}{ | m{4.4cm} | m{0.9cm}| m{5.5cm} | m{2.6cm} | } 
\hline
Contrast & DEGs & Interpretation & Results \\ 
\hline
V vs N & 43 & Genes that change expression due to virus effect regardless of diet status in bees & Table \ref{tbl:virusGenes} \\ 
\hline
NC vs NR & 941 & Genes that change expression due to diet effect in uninfected bees & Supplementary tables \ref{tbl:NCNRNCPathways} and \ref{tbl:NCNRNRPathways} \\ 
\hline
VC vs VR & 376 & Genes that change expression due to diet effect in infected bees & Supplementary tables \ref{tbl:VCVRVCPathways} and \ref{tbl:VCVRVRPathways} \\ 
\hline
VC upregulated in VC vs VR overlapped with NC upregulated in NC vs NR & 122 & ``Resilience'' genes that are turned on by good diet regardless of virus infection status in bees & Figure \ref{fig:revigo}A \\
\hline
VC upregulated in VC vs VR but NC is not upregulated in NC vs NR & 125 & ``Resistance'' genes that are turned on by good diet only in infected bees & Figure \ref{fig:revigo}B \\
\hline
\end{tabular}
\caption{Contrasts in our study for assessing GO and pathways analysis.}
  \label{tbl:contrasts}
\end{table}

@@

Mortality rates of honeybees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df=3, 55; F=10.07; p<2.18e-05). The effect of virus treatment (mixed model ANOVA, df=1, 55; F=24.343; p<7.84e-06) and diet treatment (mixed model ANOVA, df=1, 55; F=5.796; p<0.0194) were significant, but the interaction between the two factors (mixed model ANOVA, df=1, 55; F=0.062, p=0.8039) was not significant. The virus treatment was significant: For a given diet, honeybees exposed to the virus showed significantly higher mortality rate than honeybees not exposed to the virus (Tukey HSD, p<0.05). In comparing mortality levels based on pairwise comparisons, we found that bees fed Rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls. However, bees fed Chestnut pollen had no significant difference in mortality between virus infected and control groups. These results suggest the high quality Chestnut diet can ``rescue'' virus induced mortality (Figure \ref{fig:mortality}A).

IAPV titers of honeybees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df=3, 34; F=6.096; p<0.00196). The effect of virus treatment (mixed model ANOVA, df=1, 34; F=15.686; p<0.000362) was significant, but the diet treatment (mixed model ANOVA, df=1, 34; F=1.898; p>0.05) and the interaction between the two factors (mixed model ANOVA, df=1, 34; F=0.702, p>0.05) were not significant. Honeybees that were infected with the virus and fed a poor-quality Rockrose diet showed significant increases in IAPV titer volumes compared to honeybees that were not infected with the virus regardless of their diet quality (Tukey HSD, p<0.05). Overall, we interpreted this effect to mean that Rockrose pollen could not ``rescue'' high virus titers resulting from the inoculation treatment, whereas Chestnut pollen could (Figure \ref{fig:mortality}B).

@@

We observed a substantially larger number of DEGs in our diet main effect (n = 1914) than in our virus main effect (n = 43) (Supplementary table \ref{tbl:mainEffectDEGs}A and B). In the diet factor, there were more Chestnut-upregulated DEGs (n = 1033) than Rockrose-upregulated DEGs (n = 881). In the virus factor, there were more virus-upregulated DEGs (n = 38) than control-upregulated DEGs (n = 5). While these reported DEGs numbers are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Supplementary table \ref{tbl:mainEffectDEGs}A and B).

GO analysis of the Chestnut-upregulated DEGs revealed the following enriched categories (Benjamini correction < 0.05): Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Supplementary table \ref{tbl:ChestnutPathways}). GO analysis of the Rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Supplementary table \ref{tbl:RockrosePathways}).

@@

The number of DEGs across the six treatment pairings between the diet and virus factor ranged from 0 to 941 (Supplementary table \ref{tbl:pairDEGs}). Some of the trends observed in the main effect comparisons persisted: The diet level appeared to have greater influence on the number of DEGs than the virus level. Across every pair comparing the Chestnut and Rockrose levels, regardless of the virus level, the number of Chestnut-upregulated DEGs was higher than the number of Rockrose-upregulated DEGs (Supplementary table \ref{tbl:pairDEGs} C, D, E, F). For the pairs in which the diet level was controlled, the virus-exposed treatment showed equal to or more DEGs than the control treatment (Supplementary table \ref{tbl:pairDEGs} A, B). There were no DEGs between the treatment pair controlling for the control level of the virus effect (Supplementary table \ref{tbl:pairDEGs} A). These trends were observed for all three pipelines used (DESeq2, edgeR, and limma).

@@

We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset. Basic MDS plots were constructed with the DESeq2 analysis pipeline, and we could immediately determine that the Galbraith dataset may better separate the infected and uninfected honeybees better than our dataset (Supplementary figure \ref{fig:mdsPlots}). We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the MDS plots. However, through this automatically-generated plot, we can only visualize information at the sample level. Wanting to learn more about the data at the gene level, we continued with additional visualization techniques.

@@

We see that the 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays (Figure \ref{fig:pcpGalbraith}). We do see that the first replicate of the virus group appears somewhat inconsistent with the other virus replicates in Cluster 2, confirming that this trend in the data that we saw in the MDS plot carried through into the DEG results. In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure \ref{fig:pcpRutterVirus}). The replicates appear somewhat inconsistent in their esimated expression levels and there is not always such a large difference between treatment groups. We see a similar finding when we also examine a larger subset of 1,914 diet-related DEGs from our study (Supplementary figure \ref{fig:pcpRutterDiet}).

\begin{figure}[H]
\centering
% ../VirusHoneyBee/DESeq2/ClusterStandard/Clustering_data_FDR_05/
  \includegraphics[width=\textwidth]{Images/C_T_4.jpg}
  \caption{Parallel coordinate plots of the 1,019 DEGs after hiearchical clustering of size four between the virus-infected and control groups of the Galbraith study. Here ``C'' represents control, and ``T'' represents treatment of virus. Clusters 1, 3, and 4 seem to represent DEGs that were overexpressed in the virus innoculated group, and Cluster 2 seems to represent DEGs that were overexpressed in the control group. In general, the DEGs appeared as expected, but there is rather noticeable deviation of the first replicate from the virus-treated sample (``T.1'') from the other virus-treated replicates in Cluster 2. Cluster 4 also has some inconsistent replicates across the virus-treated replicates.}
  \label{fig:pcpGalbraith}
\end{figure}

\begin{figure}[H]
\centering
% /../N_V/DESeq2/ClusterStandard/Clustering_data_FDR_05
  \includegraphics[width=\textwidth]{Images/N_V_4.jpg}
  \caption{Parallel coordinate plots of the 43 DEGs after hiearchical clustering of size four between the virus-infected and control groups of our study. Here ``N'' represents non-infected control group, and ``V'' represents treatment of virus. We see from this plot that the DEG designations for this dataset do not appear as clean compared to what we saw in the Galbraith dataset in Figure \ref{fig:pcpGalbraith}.}
  \label{fig:pcpRutterVirus}
\end{figure}

We also used litre plots to examine the structure of individual DEGs: We see that indeed the individual DEGs from our data (Supplementary figure \ref{fig:litreClusterRutter}) show less consistent replications and less differences between the treatment groups compared to the individual DEGs from the Galbraith data (Supplementary figures \ref{fig:litreCluster1} and \ref{fig:litreCluster2}). For the Galbraith data, we examined individual DEGs from the first cluster (Supplementary figure \ref{fig:litreCluster1}) and second cluster (Supplementary figure \ref{fig:litreCluster2}) because the second cluster was a bit less ideal due to its inconsistent first replicate of the treatment group.

Finally, we looked at scatterplot matrices to assess the DEGs. We created standardized scatterplot matrices for each of the four clusters (Figure \ref{fig:pcpGalbraith}) of the Galbriath data (Supplementary figures \ref{fig:GalbraithClust1SM}, \ref{fig:GalbraithClust2SM}, \ref{fig:GalbraithClust3SM}, and \ref{fig:GalbraithClust4SM}). We also created standardized scatterplot matrices for our data. However, as our dataset contained 24 samples, we would need to include 276 scatterplots in our matrix, which would be too numerous to allow for efficient visual assessment of the data. As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Supplementary figures \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, and \ref{fig:RutterSM4}). We can again confirm through these plots that the DEGs from the Galbraith data appeared more as expected: Deviating more from the \textit{x=y} line in the treatment scatterplots while staying close to the \textit{x=y} line in replicate scatterplots.

Despite the virus-related DEGs (n = 1,019) from the Galbraith dataset displaying the expected patterns more than those from our dataset (n = 43), there was significant overlap (p-value < 2.2e-16) in the DEGs between the two studies (Supplementary figure \ref{fig:GRVenn}).

@@

Using the contrasts specified in Table \ref{tbl:contrasts}, we discovered 122 `resilience'' candidate genes and 125 ``resistance'' candidate genes. Within our 122 ``resilience" gene ontologies, we found functions related to metabolism (such as carbohydrate metabolism, fructose metabolism, and chitin metabolism). However, we also discovered gene ontologies related to RNA polyerase II transcription and immune response (Figure \ref{fig:revigo}A). Within our 125 ``resistance" gene ontologies, we found functions related to metabolism (such as carboyhydrate metabolism, chitin metabolism, and general metabolism) (Figure \ref{fig:revigo}B).

@@

Challenges to honey bee health are a growing concern, in particular the combined, interactive effects of nutritional stress and pathogens (Dolezal and Toth 2018). In this study, we used RNA-sequencing to probe mechanisms underlying honey bee responses to two effects, diet quality and infection with the major virus of concern, IAPV. In general, we found a major nutritional transcriptomic response, with nearly 2,000 transcripts changing in response to diet quality (rockrose/poor diet versus chestnut/good diet). The majority of these genes were upregulated in response to high quality diet, and these genes were enriched for functions (Supplmentary table \ref{tbl:ChestnutPathways}) such as nutrient signaling (insulin resistance) metabolism, and immune response (Notch signaling and JaK-STAT pathways). These data suggest high quality nutrition may allow bees to alter their metabolism, favoring investment of energy into innate immune responses.

Somewhat surprisingly, the transcriptomic response to virus infection in our experiment was fairly limited. We found only 43 transcripts to be differentially expressed, some with known immune functions (Table \ref{tbl:virusGenes}) such as argonaute-2 and a gene with similarity to MD-2 lipid recognition protein, as well as additional genes related to transcriptional regulation, and muscle contraction. The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figure \ref{fig:pcpRutterVirus} and Supplementary figures \ref{fig:mdsPlots}B, \ref{fig:litreClusterRutter}, \ref{fig:RutterSM1}, \ref{fig:RutterSM2}, \ref{fig:RutterSM3}, and \ref{fig:RutterSM4}).

Given the noisy nature of our data, and our desire to hone in on genes with real expression differences, we compared our data to the Galbraith study (\citealt{galbraith}), which also examined bees response to viral infection. In contrast to our study, Galbraith et al. identified a large number of virus responsive trancripts, and generally had less noise in their data (Figure \ref{fig:pcpGalbraith} and Supplementary figures \ref{fig:mdsPlots}A, \ref{fig:litreCluster1}, \ref{fig:litreCluster2}, \ref{fig:GalbraithClust1SM}, \ref{fig:GalbraithClust2SM}, \ref{fig:GalbraithClust3SM}, and \ref{fig:GalbraithClust4SM}). To identify the most reliable virus-responsive genes from our study, we looked for overlap in the DEGs associated with virus infection on both experiments. We found a large, statistically significant (p-value < 2.2e-16) overlap, with 26/38 (68\%) of virus-responsive DEGs from our study also showing response to virus infection in Galbraith et al. (Supplementary figure \ref{fig:GRVenn}). This result gives us confidence that, although noisy, we were able to uncover consistent, replicable gene expression responses to virus infection with our data.

Data visualization is a useful method to identify noise and robustness in RNA-seq data. In this study, we used extensive data visualization to improve the interpretation of our RNA-seq results.  For example, the DESeq2 package comes with certain visualization options that are popular in RNA-sequencing analysis. One of these visualization is the multidimensional scaling (MDS) plot, which allows users to visualize the similarity between samples within a dataset. We could determine from this plot that indeed the Galbraith data may show more similarity between its replicates and differences between its treatments compared to our data (Supplementary figure \ref{fig:mdsPlots}). However, the MDS plot only shows us information at the sample level. We wanted to investigate how these differences in the signal:to:noise ratios of the datasets would affect the structure of any resulting DEGs. As a result, we also used three plotting techniques from the bigPint package: We investigated the 1,019 virus-related DEGs from the Galbraith dataset and the 43 virus-related DEGs from our dataset using parallel coordinate lines, litre plots, and scatterplot matrices. To prevent overlapping issues in our plots, we used a hierarchical clustering technique for the parallel coordinate lines to separate the set of DEGs into smaller groups. We also needed to examine four subsets of samples from the Galbraith dataset to make effective use of the scatterplot matrices. After these tailorizations, we determined that the same patterns we saw in the MDS plots regarding the entire dataset extended down the pipeline analysis into the DEG calls: Even the DEGs from the Galbraith datasaet showed more similarity between their replicates and differences between their treatments compared to those from our data. However, the 365 DEGs from the Galbraith data in Cluster 2 of Figure \ref{fig:pcpGalbraith} showed an inconsistent first replicate in the treatment group (``T.1''), which was something we observed in the MDS plot. This indicates that this feature also extended down the analysis pipeline into DEG calls. We believe these visualization applications can be useful for future researchers analyzing RNAs-sequencing data to quickly and effectively ensure that the DEG calls look reliable or at least overlap with DEG calls from similar studies that look reliable. We believe this type of visualization exploration can be especially crucial when studying complex organisms that do not have genetic identicalness or similarity between replicates and/or when using experiments that may lack rigid design control.

One of the goals of this study was to use our RNA-seq data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of ``resistance'' or ``resilience''. We attempted to address this question through specific gene expression contrasts (Table \ref{tbl:contrasts}), accompanied by GO analysis of the associated gene lists. We found an approximately equal number of resistance (n = 125) and resilience (n = 122) related candidate genes, suggesting both processes may be playing significant roles in dietary buffering from pathogen infection. Resilience candidate genes had functions related to carbohydrate metabolism, chitin metabolism, immune response, and regulation of transcription. Resistance candidate genes had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion.

Overall, these data suggest complex transcriptomic responses to multiple stressors in honey bees. Diet has the potential for large and profound effects on transcriptional responses in honey bees, and differences in diet may set up the potential for both resistance and resilience to virus infection. Moreover, this study in general also demonstrated the possible benefits of using data visualizations and multiple datasets to address inherently messy biological data. For instance, by verifying the subsantial overlap in our DEG lists to those obtained in another study that addressed a similar question but in a more controlled manner, we were able to place much higher confidence in the differential gene expression results from our otherwise noisy data. We hope these results underline the need for researchers to use data visualization techniques to understand and interpret RNA-sequencing datasets.

\section{Appendix}

\begin{table}[H]
  \includegraphics[width=0.65\textwidth]{Images/mainEffectDEGs}
  \caption{Number of DEGs across three analysis pipelines for (A) the diet effect in our study, (B) the virus main effect in our study, and (C) the virus main effect in the Galbraith study. For the diet effects, ``C'' represents Chestnut diet and ``R'' represents Rockrose diet. For the virus effects, ``V'' represents virus-innoculated and ``C'' represents control non-innoculated. Green cells represent the level that showed a larger number of DEGs.}
  \label{tbl:mainEffectDEGs}
\end{table}
